---
title: "Lab 14 Genome Informatics with Homework"
format: pdf
---

##Introduction to Genome Informatics Lab
#Section 1: Identify genetic variants of interest

Q5: Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are homozygous for the asthma associated SNP (G|G)?

Downloaded CSV file from Ensemnble < https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=105535077

Read CSV file and analyze the sample:

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

A5:
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```
14% are homozygous for asthma within MXL population sample in LA


Q6. Back on the ENSEMBLE page, use the “search for a sample” field above to find the particular sample HG00109. This is a male from the GBR population group. What is the genotype for this sample?

Now let's look at diff population, specifically GBR, "Great Britain"

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G
```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2 )
```

A: 29.67% are homozygous for asthma within GBR population sample in Great Britain 

#Section 4: Population Scale Analysis
How many samples doe we have?
```{r}
expr <- read.table(("rs8067378_ENSG00000172057.6.txt"))
head(expr)
```
Total individuals?
```{r}
nrow(expr)
```

How many of each genotype?
```{r}
table(expr$geno)
```

How do we deplay these results?
```{r}
library(ggplot2)
```

let's make a boxplot with this data:
```{r}
ggplot(expr) + aes(x=geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```
Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

A14: According to the boxplot above, epression of ORMDL3 is highest when individuals experess homologous alleles A|A, and lowest for G|G. The SNP does effect the expression of ORDML3
